Memes: an R interface to the MEME Suite

### 2020-10-27 manuscript.Rmd
reorganizing the results.

The core functionality of `memes` revolves around the interface to MEME Suite
tools. Each supported MEME Suite tool is assigned a `run` function
(`runDreme()`, `runMeme()`, `runAme()`, `runFimo()`, `runTomTom()`), which
accept sequence and motif inputs as needed and any valid commandline flags as
additional function parameters. A full list of currently supported MEME Suite
tools and their function is found in \ref{tab:tool_function}.

### Data Structures

MEME Suite tools are run on the commandline and use files stored on-disk as
input while returning a series of output files containing varying data types.
`memes` reduces each MEME Suite tool to an R function which internally handles
file I/O such that all MEME Suite calling functions accept R objects as input,
and return R objects as output. Sequence inputs are stored in `BioStrings`
format, an R/Bioconductor package for storing biopolymer sequence data
\cite{BioStrings}. Motif inputs are passed as `universalmotif` objects, another
R/Bioconductor package for representing motif matrices and their associated
metadata \cite{universalmotif}. `memes` will also accept paths to files
on disk, such as fasta files for sequence inputs, and meme format files for
motif inputs, reducing the need to read large files into memory.

**Restructure below:**
 - MEME tools return HTML reports for each tool which display data in user-friendly way, but poor for data processing. However, some of the simple raw data (like tsv data) don't contain the full set of statistics reported to the user. These data are contained in XML files, which are difficult to parse.

Although MEME Suite tools accept consistent data type inputs, they do not return
consistent data structures across tools. For example, the TomTom tool, which
compares query motifs to a database of known motifs, returns tab-separated
results that do not contain the frequency matrix representations of the matched
motifs. Instead, users must write custom code to parse these matrices back out
from the input databases, creating additional friction for creating novel data
visualizations and exploratory analyses. Often, output data are also returned in
XML format, which although they hold all relevant information, are difficult to
parse. To our knowledge, no publicly available tools exist to parse MEME Suite
XML formatted data, therefore users must write custom code to extract these
data. `memes` exports data import functions for each supported MEME Suite tool,
which import these data as R data.frames. These functions also work internally to
create the seamless file I/O when calling MEME Suite tools, but the import
functions can be called directly by users to import previously generated data
for use in R.

 - **NOTE: Biopython has a MEME/MAST parser but it works on tsv https://biopython.org/docs/1.74/api/Bio.motifs.meme.html**
 - **Actually, they merged XML import in March: https://github.com/biopython/biopython/pull/2354**
 - **Need to revise above statement, but note that most tools don't have support**
 - ? To our knowledge, no publicly available tools exist to parse XML formatted data from all MEME Suite tools ...

Many of the MEME Suite tools return files which
contianing descriptive properties of motifs. For example, MEME and DREME return de-novo
discovered motifs from query sequences along with statistical information about
their enrichment. The `universalmotif` data structure is a powerful
R/Bioconductor representation for motif matrices and their associated metadata \cite{universalmotif}. 
However, `universalmotif` objects are not ideal for performing bulk manipulation
or queries of metadata, nor do they support complex data structures such as
nested elements. Many MEME Suite analysis pipelines involve identifying motifs
which are further analyzed by multiple tools, creating additional metadata which
cannot be easily tracked in bulk using `universalmotif` format. Thus
an ideal representation for these data is one which can hold an unlimited amount
of metadata, contain complex hierarchies, and be easily manipulated using
standard analysis tools. `memes` therefore extends the `universalmotif` format
by providing data.frames holding motif metadata which retain links to their
`universalmotif` representation. These universalmotif
data.frames combine the advantages of the `universalmotif` framework for motif
manipulation with the extensibility and flexibility of R data.frames. These
`universalmotif` data.frames form the basis for a majority of `memes` data
outputs. These structures are also valid input types to `memes` functions, and
when used as such, output data are appended as new columns to the input data,
ensuring data provenance.

>>>>

MEME Suite tools are run on the commandline and use files stored on-disk as
input while returning a series of output files containing varying data types. 
As a wrapper of MEME tools, `memes` functions similarly by assigning each
supported MEME Suite tool to a `run` function (`runDreme()`,
`runMeme()`, `runAme()`, `runFimo()`, `runTomTom()`), which
internally writes input objects to files on disk, runs the tool, then imports
the data as R objects. These functions accept sequence and motif inputs as
required by the tool. Sequence inputs are accepted in `BioStrings` format, an
R/Bioconductor package for storing biopolymer sequence data \cite{BioStrings}.
Motif inputs are passed as `universalmotif` objects, another R/Bioconductor
package for representing motif matrices and their associated metadata
\cite{universalmotif}. `memes` `run` functions will also accept paths to files
on disk, such as fasta files for sequence inputs, and meme format files for
motif inputs, reducing the need to read large files into memory. Finally, each
`run` function contains optional function parameters mirroring the commandline
arguments provided by each MEME Suite tool. In this way, `memes` provides a
feature-complete interface to the supported MEME Suite tools.

### Output Data Structures

MEME tools return HTML reports for each tool that display data in a
user-friendly way, however, these files are not ideal for downstream processing.
Depending on the tool, data are also returned in tab-separated, or XML format,
which are more amenable to computational processing. However, in the case that
tab-separated data are returned, results are often incomplete. For example, the
TomTom tool, which compares query motifs to a database of known motifs, returns
tab-separated results that do not contain the frequency matrix representations
of the matched motifs. Instead, users must write custom code to parse these
matrices back out from the input databases, creating additional barriers to
analysis. In the case that data are returned in XML format, these files contain
all relevant data, however are difficult to parse, again requiring users to
write custom code to extract the necessary data. Finally, the data types
contained in MEME Suite outputs are complex and multidimensional, and thus
require special data structures to properly organize the data in memory.

`memes` provides data import functions for each supported MEME Suite tool, which
import these data as modified R data.frames (described in detail below). These
functions can be called directly by users to import data previously generated by
the commandline or webserver versions of the MEME Suite for use in R. These
import functions also underlie the import step internal to each of the `run`
functions.

### structured data.frames hold multidimensional output data

 - as mentioned previously, the multidimensional naature of meme results is difficult to represent
 - Output data usually contain motifs & descriptive statistics about those motifs
 - motif info can be stored in universalmotif format, however, not enough space for data
 - data.frames do good job of storing data (incl nested data), but cannot store motifs
 - we solve this problem by storing the unviversalmotif inside the data.frame (which is valid R)
 - we add nested list columns as needed to link multidimensional data
 - advantage of this approach over an S4 is all normal tools work on it (tidyverse), joins, etc.

Many of the MEME Suite tools return files contianing descriptive
properties of motifs. For example, MEME and DREME return de-novo discovered
motifs from query sequences along with statistical information about their
enrichment. The `universalmotif` data structure is a powerful R/Bioconductor
representation for motif matrices and their associated metadata
\cite{universalmotif}.  However, `universalmotif` objects are not ideal for
performing bulk manipulation or queries of metadata, nor do they support complex
data structures such as nested elements. Many MEME Suite analysis pipelines
involve identifying motifs which are further analyzed by multiple tools,
creating additional metadata which cannot be easily tracked in bulk using
`universalmotif` format. Thus an ideal representation for these data is one
which can hold an unlimited amount of metadata, contain complex hierarchies, and
be easily manipulated using standard analysis tools. `memes` therefore extends
the `universalmotif` format by providing data.frames holding motif metadata
which retain links to their `universalmotif` representation. These
universalmotif data.frames combine the advantages of the `universalmotif`
framework for motif manipulation with the extensibility and flexibility of R
data.frames. These `universalmotif` data.frames form the basis for a majority of
`memes` data outputs. These structures are also valid input types to `memes`
functions, and when used as such, output data are appended as new columns to the
input data, ensuring data provenance.
